White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-12-15 papers/wdmerger_I/paper.tex
Fix IACS acknowledgement
Add affiliations

\author{Max P. Katz, Michael Zingale, Alan C. Calder, F. Douglas Swesty, Ann S. Almgren, Weiqun Zhang}
%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possible progenitor that has 
merited recent theoretical attention is the white dwarf merger scenario.
This has the potential to naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been relatively few fully self-consistent simulations 
of merging binary white dwarf systems using mesh-based hydrodynamics. 
This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
adaptive mesh refinement. In this paper we describe our numerical 
methodology and discuss our implementation in the compressible hydrodynamics code \castro.
For merging white dwarf simulations we use \castro's capability to solve the Euler equations, and
the Poisson equation for self-gravity, and couple the gravitational force to the hydrodynamics.
Standard techniques for coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system for our problem, 
but recent advances in the literature allow progress on this 
issue and we discuss our implementation here. We present a
set of test problems demonstrating the extent to which our software sufficiently
models a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

This research has made use of the AstroBetter blog and wiki. 
This research has made use of NASA's Astrophysics Data System 
Bibliographic Services.

This research was supported by NSF award AST-1211563. An
award of computer time was provided by the Innovative and Novel
Computational Impact on Theory and Experiment (INCITE) program.  This
research used resources of the Oak Ridge Leadership Computing Facility
located in the Oak Ridge National Laboratory, which is supported by
the Office of Science of the Department of Energy under Contract
DE-AC05-00OR22725. Projects AST006 and AST106 supported use of the ORNL/Titan resource. 
This research is part of the Blue Waters sustained-petascale computing project, 
which is supported by the National Science Foundation (awards OCI-0725070 
and ACI-1238993) and the state of Illinois. Blue Waters is a joint 
effort of the University of Illinois at Urbana-Champaign and its 
National Center for Supercomputing Applications.
This research used resources of the National Energy Research Scientific Computing
Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No. DE-AC02-05CH11231.  
This work used the Extreme Science and Engineering Discovery Environment (XSEDE), 
which is supported by National Science Foundation grant number ACI-1053575. 
Project AST100037 supported use of the resources NICS/Kraken and NICS/Darter.

>>>>

\author{Max P. Katz\altaffilmark{1}}
\author{Michael Zingale\altaffilmark{1}}
\author{Alan C. Calder\altaffilmark{1,2}}
\author{F. Douglas Swesty\altaffilmark{1}}
\author{Ann S. Almgren\altaffilmark{3}}
\author{Weiqun Zhang\altaffilmark{3}}

\altaffiltext{1}
{
  Department of Physics and Astronomy,
  Stony Brook University, Stony Brook, NY, 11794-3800, USA
}

\altaffiltext{2}
{
  Institute for Advanced Computational Sciences,
  Stony Brook University, Stony Brook, NY, 11794-5250, USA
}

\altaffiltext{3}
{
  Center for Computational Sciences and Engineering,
  Lawrence Berkeley National Laboratory, Berkeley, CA 94720
}

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
The Type Ia supernova progenitor problem is one of the most perplexing and 
exciting problems in astrophysics, requiring detailed numerical modeling to 
complement the observations of these explosions. One possible progenitor that has 
merited recent theoretical attention is the white dwarf merger scenario.
This has the potential to naturally explain many of the observed characteristics of 
Type Ia supernovae, as well as their distribution in time and space.
To date there have been relatively few fully self-consistent simulations 
of merging binary white dwarf systems using mesh-based hydrodynamics. 
This is the first paper in a series designed to 
describe simulations of these systems using a mesh-based hydrodynamics code with 
adaptive mesh refinement. In this paper we describe our numerical 
methodology and discuss our implementation in the compressible hydrodynamics code \castro.
For merging white dwarf simulations we use \castro's capability to solve the Euler equations, and
the Poisson equation for self-gravity, and couple the gravitational force to the hydrodynamics.
Standard techniques for coupling gravitation and rotation forces to the hydrodynamics do 
not do an adequate job conserving the total energy of the system for our problem, 
but recent advances in the literature allow progress on this 
issue and we discuss our implementation here. We present a
set of test problems demonstrating the extent to which our software sufficiently
models a system where large amounts of mass are advected on the computational 
domain over very long timescales. Future papers in this series will describe
our treatment of the initial conditions of these systems and will 
examine the early phases of the merger to determine its viability
for triggering a thermonuclear detonation.

@@

This research was supported by NSF award AST-1211563. An
award of computer time was provided by the Innovative and Novel
Computational Impact on Theory and Experiment (INCITE) program.  This
research used resources of the Oak Ridge Leadership Computing Facility
located in the Oak Ridge National Laboratory, which is supported by
the Office of Science of the Department of Energy under Contract
DE-AC05-00OR22725. Projects AST006 and AST106 supported use of the ORNL/Titan resource. 
This research is part of the Blue Waters sustained-petascale computing project, 
which is supported by the National Science Foundation (awards OCI-0725070 
and ACI-1238993) and the state of Illinois. Blue Waters is a joint 
effort of the University of Illinois at Urbana-Champaign and its 
National Center for Supercomputing Applications.
This research used resources of the National Energy Research Scientific Computing
Center, which is supported by the Office of Science of the
U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
Results in this paper were obtained using the high-performance
computing system at the Institute for Advanced Computational Science
at Stony Brook University.
This work used the Extreme Science and Engineering Discovery Environment (XSEDE), 
which is supported by National Science Foundation grant number ACI-1053575. 
Project AST100037 supported use of the resources NICS/Kraken and NICS/Darter.

This research has made use of NASA's Astrophysics Data System 
Bibliographic Services. In addition, this research has made use
of the AstroBetter blog and wiki.
